Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-05 paper.qmd
Added mse plot, aic plot for pbmc need to find to remove duplicates of mse_df

```{r}
#| warning: false
#| echo: false
library(dplyr)
# remotes::install_github("jlmelville/snedata")
library(snedata)
library(ggflowchart)
library(purrr) ## map function
library(gridExtra) ## for grid.arrange
library(rsample)
library(DT)
library(ggbeeswarm)
library(ggplot2)
library(readr)

library(Rtsne)
library(umap)
library(phateR)
library(patchwork)
library(langevitour)
library(colorspace)
library(kableExtra)

library(grid)

# ## For paralleization
# library(foreach)
# library(doParallel)

# library(xtable)
# 
# xtable2kable <- function(x) {
#   out <- capture.output(print(x, table.placement = NULL))[-(1:2)]
#   out <- paste(out, collapse = "\n")
#   structure(out, format = "latex", class = "knitr_kable")
# }

# install.packages("Seurat")
# remotes::install_github("satijalab/seurat-data")
#library(SeuratData) ## For the application

# use_python("~/miniforge3/envs/pcamp_env/bin/python")
# use_condaenv("pcamp_env")
# 
# reticulate::source_python(paste0(here::here(), "/examples_for_the_paper/function_scripts/Fit_PacMAP_code.py"))
# reticulate::source_python(paste0(here::here(), "/examples_for_the_paper/function_scripts/Fit_TriMAP_code.py"))

set.seed(20240110)

source("quollr_code.R", local = TRUE)
source("nldr_code.R", local = TRUE)
```

@@

# data <- data |>
#   mutate(ID = row_number())
# 
# data_split <- initial_split(data)
# training_data <- training(data_split) |>
#   arrange(ID)
# test_data <- testing(data_split) |>
#   arrange(ID)
training_data <- read_rds("data/s_curve/s_curve_training.rds")
test_data <- read_rds("data/s_curve/s_curve_test.rds")
```

@@

#plot_list <- list()

tSNE_data <- read_rds("data/s_curve/s_curve_tsne_27.rds")
#(perplexity: ", calculate_effective_perplexity(data), ")
plot_list1 <- plot_tSNE_2D(tSNE_data) + #ggtitle("(a)") + 
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

#(n-neighbors: 50)
plot_list2 <- plot_UMAP_2D(UMAP_data) + #ggtitle("(b)") + 
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

plot_list3 <- plot_PHATE_2D(PHATE_data) + #ggtitle("(c)") + 
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

#(n-inliers: 5, \n n-outliers: 4, n-random: 3)
plot_list4 <- plot_TriMAP_2D(TriMAP_data) + #ggtitle("(d)") + 
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'd', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

@@

#(knn: 10, init: random, \n MN-ratio: 0.9, FP-ratio: 2)
plot_list5 <- plot_PaCMAP_2D(PaCMAP_data) + #ggtitle("(e)") + 
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'e', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

@@

@@

**(a) Determine the number of bins along the x-axis** ($b_1$)

@@

$$
 \text{A} = \frac{\sqrt{3}}{2}h^2
$$ {#eq-equation3}

To construct regular hexagons, $A = 1$ (see @fig-binsize) use as the default setting. Then, the diameter ($h$) of the regular hexagons is calculated (see @eq-equation4).

$$
  \text{h} = \sqrt{\frac{2}{\sqrt{3}}A}
$$ {#eq-equation4}

@Carr2013 mentioned about the relationship between the diameter ($h$) of regular hexagons and the height ($y$) of the plotting region. According to our algorithm, the height ($y$) of the plotting region is the the range of 2D embedding component 1 ($r_1$) (see @eq-equation6).

$$
h = \frac{r_1}{b_1}
$$ {#eq-equation6}

After rearranging the @eq-equation6 as @eq-equation5, $b_1$ is computed. The $b_1$ value is rounded up to the nearest whole number to have an integer value.

$$
b_1 = \frac{r_1}{h}
$$ {#eq-equation5}

**(b) Determine the shape parameter** ($s$)

In this step, we determine the shape parameter ($s$) for the hexagonal bins, which significantly influences their shape and arrangement within the grid. The $s$ in the hexagonal binning algorithm is defined as the ratio of the height ($y$) to the width ($x$) of the plotting region as defined in @eq-equation1 [@Carr2013].

$$
s = \frac{y}{x}
$$ {#eq-equation1}

The shape parameter ($s$) of our algorithm is calculated as the ratio of the ranges of 2D embedding components, where $r_1$ and $r_2$ represent the ranges of 2D embedding components 1 and component 2, respectively (see @eq-equation2).

$$
s = \frac{r_2}{r_1}
$$ {#eq-equation2}

**(c) Determine the number of bins along the y-axis** ($b_2$)

Next, the number of bins along the y-axis is computed based on the number of bins along the x-axis ($b_1$) and the shape parameter ($s$) (see @eq-equation12) [@Carr2013].

$$
b_2 = 2 * \left(\frac{(b_1 \times s)}{\sqrt(3)} + 1.5001 \right)
$${#eq-equation12}

**(d) Determine the total number of bins** ($b$)

The total number of bins is determined by multiplying the number of bins along the x-axis ($b_1$) with the number of bins along the y-axis ($b_2$) (see @eq-equation13).

$$
b = b_1 \times b_2
$${#eq-equation13}

**Step 2: Obtain bin centroids**

@@

# UMAP_fit <- umap(training_data |> dplyr::select(-ID), n_neighbors = 50, n_components =  2)
# 
# UMAP_data <- UMAP_fit$layout |>
#   as.data.frame()
# names(UMAP_data)[1:(ncol(UMAP_data))] <- paste0(rep("UMAP",(ncol(UMAP_data))), 1:(ncol(UMAP_data)))
# 
# UMAP_data <- UMAP_data |>
#   mutate(ID = training_data$ID)

@@

plot_list1_gau <- plot_tSNE_2D(tSNE_data_gau) + #ggtitle("(a)") + 
  geom_point(size = 0.00001, colour = "#e41a1c") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

plot_list2_gau <- plot_UMAP_2D(UMAP_data_gau) + #ggtitle("(b)") +
  geom_point(size = 0.00001, colour = "#377eb8") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

plot_list3_gau <- plot_PHATE_2D(PHATE_data_gau) + #ggtitle("(c)") +
  geom_point(size = 0.00001, colour = "#4daf4a") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

plot_list4_gau <- plot_TriMAP_2D(TriMAP_data_gau) + #ggtitle("(d)") +
  geom_point(size = 0.00001, colour = "#984ea3") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'd', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

plot_list5_gau <- plot_PaCMAP_2D(PaCMAP_data_gau) + #ggtitle("(e)") +
  geom_point(size = 0.00001, colour = "#ff7f00") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'e', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)
```

@@

## Import data
#df_2 <- read_rds("data/pbmc/pbmc.rds")
training_data_pbmc <- read_rds("data/pbmc/pbmc_3k_festem/pbmc_pca_50.rds")
training_data_pbmc <- training_data_pbmc[, 1:15] |>
  mutate(ID = 1:NROW(training_data_pbmc))
#test_1 <- read_rds("data/pbmc/pbmc_test.rds")

#tSNE_pbmc <- read_rds("data/pbmc/pbmc_tsne_30.rds")
UMAP_pbmc <- read_rds("data/pbmc/pbmc_3k_festem/pbmc_umap.rds")
# PHATE_pbmc <- read_rds("data/pbmc/pbmc_phate.rds")
# TriMAP_pbmc <- read_rds("data/pbmc/pbmc_trimap.rds")
# PaCMAP_pbmc <- read_rds("data/pbmc/pbmc_pacmap.rds")

#load("data/pbmc/pbmc_3k_festem/pbmc3k_label.RData")

# plot_list1_pbmc <- plot_tSNE_2D(tSNE_pbmc) + #ggtitle("(a)") + 
#   theme_linedraw() +
#   theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
#         axis.title.x = element_blank(), axis.title.y = element_blank(),
#         axis.text.x = element_blank(), axis.ticks.x = element_blank(),
#         axis.text.y = element_blank(), axis.ticks.y = element_blank(),
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
#   annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

plot_list2_pbmc <- UMAP_pbmc |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = cell_label))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#cab2d6", "#ccebc5", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#ffed6f", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#b2df8a", "#bc80bd", "#33a02c", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  geom_text(aes(x=UMAP1,y = UMAP2,label = cell_label), data = class_avg,inherit.aes = F, color = "black",fontface = "bold",size = 2) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) #+
  # guides(color = guide_legend(title = "Cell type"),
  #        color = guide_legend(nrow = 3)) #+
  # annotate(geom = 'text', label = 'b', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

# plot_list3_pbmc <- plot_PHATE_2D(PHATE_pbmc) + #ggtitle("(c)") + 
#   theme_linedraw() +
#   theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
#         axis.title.x = element_blank(), axis.title.y = element_blank(),
#         axis.text.x = element_blank(), axis.ticks.x = element_blank(),
#         axis.text.y = element_blank(), axis.ticks.y = element_blank(),
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
#   annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

# plot_list4_pbmc <- plot_TriMAP_2D(TriMAP_pbmc) + #ggtitle("(d)") + 
#   theme_linedraw() +
#   theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
#         axis.title.x = element_blank(), axis.title.y = element_blank(),
#         axis.text.x = element_blank(), axis.ticks.x = element_blank(),
#         axis.text.y = element_blank(), axis.ticks.y = element_blank(),
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
#   annotate(geom = 'text', label = 'd', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)
# 
# plot_list5_pbmc <- plot_PaCMAP_2D(PaCMAP_pbmc) + #ggtitle("(e)") + 
#   theme_linedraw() +
#   theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
#         axis.title.x = element_blank(), axis.title.y = element_blank(),
#         axis.text.x = element_blank(), axis.ticks.x = element_blank(),
#         axis.text.y = element_blank(), axis.ticks.y = element_blank(),
#         panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
#   annotate(geom = 'text', label = 'e', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)
```

@@

## To draw with AIC
aic_plot_pbmc <- ggplot(MSE_df_pbmc |> dplyr::filter(data_type == "training") |> dplyr::filter(method == "UMAP"), aes(x = number_of_bins,
                                                                                 y = total_error,
                                                                                 color = method
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 256, linetype="dashed", 
                color = "red", size=0.5) +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#377eb8", "#e41a1c", "#4daf4a", "#984ea3", "#ff7f00")) +
  ylab("AIC") +
  xlab("total number of bins")
## Effective number of bins along x-axis

>>>>

```{r}
#| warning: false
#| echo: false
library(quollr)
library(dplyr)
# remotes::install_github("jlmelville/snedata")
library(snedata)
library(ggflowchart)
library(purrr) ## map function
library(gridExtra) ## for grid.arrange
library(rsample)
library(DT)
library(ggbeeswarm)
library(ggplot2)
library(readr)

library(Rtsne)
library(umap)
library(phateR)
library(patchwork)
library(langevitour)
library(colorspace)
library(kableExtra)
library(grid)
set.seed(20240110)

#source("quollr_code.R", local = TRUE)
source("nldr_code.R", local = TRUE)
```

@@

training_data <- read_rds("data/s_curve/s_curve_training.rds")
test_data <- read_rds("data/s_curve/s_curve_test.rds")
```

@@

tSNE_data <- read_rds("data/s_curve/s_curve_tsne_27.rds")

plot_list1 <- plot_tSNE_2D(tSNE_data) +  
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

#(n-neighbors: 50)
plot_list2 <- plot_UMAP_2D(UMAP_data) + 
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

plot_list3 <- plot_PHATE_2D(PHATE_data) + 
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

#(n-inliers: 5, \n n-outliers: 4, n-random: 3)
plot_list4 <- plot_TriMAP_2D(TriMAP_data) + 
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'd', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

@@

#(knn: 10, init: random, \n MN-ratio: 0.9, FP-ratio: 2)
plot_list5 <- plot_PaCMAP_2D(PaCMAP_data) +  
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'e', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

@@

Then, after performing NLDR, to have regular hexagons, 2D embedding data are scaled accordingly.

@@

**(a) Determine the number of bins along the x axis** ($b_1$)

@@

In the process of computing hexagonal bin configurations, the first step involves determining the number of bins along the x-axis. This begins with the initialization of parameters such as the hexagonal size ($s$) and the buffer along the x-axis. Subsequently, the range of the first 2D embedding component ($r_1$) is calculated to ensure comprehensive coverage of the data. This range is then adjusted by incorporating the buffer amount, effectively expanding the boundary to accommodate potential outliers or edge cases. The default buffer amount along the x-axis is $\sqrt{3} * s * 1.5$. Next, the horizontal spacing ($h$) between hexagons is computed based on the hexagon size. Finally, the number of bins along the x-axis is computed based on the adjusted range ($r_1 + buffer_{x}$) and the horizontal spacing ($h$).

$$
 h = \sqrt{3} * s
$$ {#eq-equation3}

$$
 b_1 = \frac{r_1 + buffer_{x}}{h}
$$ {#eq-equation4}

**(b) Determine the number of bins along the y-axis** ($b_2$)

Next, the number of bins along the y-axis is computed based on the number of bins along the x-axis ($b_1$) and the shape parameter ($s$) (see @eq-equation4) [@Carr2013].

$$
  v = 1.5 * s
$$ {#eq-equation5}

$$
 b_2 = \frac{r_2 + buffer_{y}}{v}
$$ {#eq-equation6}

**(c) Determine the total number of bins** ($b$)

The total number of bins is determined by multiplying the number of bins along the x-axis ($b_1$) with the number of bins along the y-axis ($b_2$) (see @eq-equation5).

$$
b = b_1 \times b_2
$${#eq-equation5}

**Step 2: Obtain bin centroids or bin means**

@@

@@

plot_list1_gau <- plot_tSNE_2D(tSNE_data_gau) + 
  geom_point(size = 0.00001, colour = "#e41a1c") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

plot_list2_gau <- plot_UMAP_2D(UMAP_data_gau) + 
  geom_point(size = 0.00001, colour = "#377eb8") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

plot_list3_gau <- plot_PHATE_2D(PHATE_data_gau) + 
  geom_point(size = 0.00001, colour = "#4daf4a") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

plot_list4_gau <- plot_TriMAP_2D(TriMAP_data_gau) + 
  geom_point(size = 0.00001, colour = "#984ea3") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'd', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

plot_list5_gau <- plot_PaCMAP_2D(PaCMAP_data_gau) + 
  geom_point(size = 0.00001, colour = "#ff7f00") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5), 
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'e', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)
```

@@

## Import data
training_data_pbmc <- read_rds("data/pbmc/pbmc_3k_festem/pbmc_pca_50.rds")
training_data_pbmc <- training_data_pbmc[, 1:15] |>
  mutate(ID = 1:NROW(training_data_pbmc))

umap_pbmc <- read_rds("data/pbmc/pbmc_3k_festem/pbmc_umap.rds")
umap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = umap_pbmc, 
                                    x = "UMAP1", y = "UMAP2"))) |>
  dplyr::mutate(cell_label = umap_pbmc$cell_label) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

class_avg <- umap_pbmc_scaled |>
  group_by(cell_label) |>
  summarise(
    UMAP1 = median(UMAP1),
    UMAP2 = median(UMAP2)
  ) |>
  mutate(cell_label = as.numeric(cell_label)) |>
  arrange(cell_label)

@@

plot_list2_pbmc <- umap_pbmc_scaled |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = cell_label))+
    geom_point(alpha=0.5) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(legend.position = "none", plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#b15928", "#1f78b4", "#cab2d6", "#ccebc5", "#fb9a99", "#e31a1c", "#6a3d9a", "#ff7f00", "#ffed6f", "#fdbf6f", "#ffff99", "#a6cee3", "#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3", "#fdb462", "#b3de69", "#fccde5", "#d9d9d9", "#b2df8a", "#bc80bd", "#33a02c", "#ccebc5", "#ffed6f", "#000000", "#bdbdbd")) +
  geom_text(aes(x=UMAP1,y = UMAP2,label = cell_label), data = class_avg,inherit.aes = F, color = "black",fontface = "bold",size = 2) +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) 
```

```{r}
#| warning: false
#| echo: false
#| message: false

#tot_num_bin_vec <- 2:51
hex_size_vec <- seq(0.02, 2, by = 0.01)

vec <- stats::setNames(rep("", 6), c("num_bins", "aic", "mse", "num_bins_x", "num_bins_y", "hex_size"))  ## Define column names

mse_df_pbmc <- dplyr::bind_rows(vec)[0, ]
mse_df_pbmc <- mse_df_pbmc |>
  dplyr::mutate_if(is.character, as.numeric)

for (i in 1:length(hex_size_vec)) {

  num_bin_list <- calc_bins(data = umap_pbmc_scaled, 
            x = "UMAP1", y = "UMAP2", 
            hex_size = hex_size_vec[i], buffer_x = NA, buffer_y = NA)

  num_bins_x <- num_bin_list$num_x
  num_bins_y <- num_bin_list$num_y

  model_object <- fit_highd_model( training_data = training_data_pbmc, 
                                   nldr_df_with_id = umap_pbmc_scaled |> dplyr::select(-cell_label), 
                                   x = "UMAP1", y = "UMAP2", 
                                   num_bins_x = num_bins_x, 
                                   num_bins_y = num_bins_y, 
                                   x_start = NA, y_start = NA, 
                                   buffer_x = NA, buffer_y = NA, 
                                   hex_size = hex_size_vec[i],
                                   is_rm_lwd_hex = FALSE, 
                                   benchmark_to_rm_lwd_hex = NA, 
                                   col_start_2d = "UMAP", 
                                   col_start_highd = "PC_")

  pred_emb_list <- predict_emb(test_data = training_data_pbmc, 
                                  df_bin_centroids = centroid_df_training, 
                                  df_bin = avg_df_training, type_NLDR = "UMAP")

  pred_df_training <- as.data.frame(do.call(cbind, pred_emb_list))

  eval_list <- gen_summary(test_data = training_data_pbmc, 
                                  prediction_df = pred_df_training, 
                                  df_bin = avg_df_training, col_start = "PC_")

  mse_df_pbmc <- mse_df_pbmc |>
    tibble::add_row(num_bins = num_bins_x * num_bins_y,
                    aic = eval_list$aic,
                    mse = eval_list$mse,
                    num_bins_x = num_bins_x,
                    num_bins_y = num_bins_y,
                    hex_size = hex_size_vec[i])

## If same total number of bins ocurred only select ones with minimum error
duplicate_df_list <- mse_df_pbmc[duplicated(mse_df_pbmc$num_bins), ] |>
  dplyr::arrange(num_bins) |>
  dplyr::group_split(num_bins)

duplicate_df <- data.frame(matrix(nrow = 0, ncol = 0))

for (i in 1:length(duplicate_df_list)) {

  dd <- duplicate_df_list[[i]] |>
    dplyr::filter(mse == min(duplicate_df_list[[i]]$mse))

  duplicate_df <- dplyr::bind_rows(duplicate_df, dd)

mse_df_pbmc <- dplyr::bind_rows(mse_df_pbmc[!duplicated(mse_df_pbmc$num_bins), ], duplicate_df)
```

@@

## To draw with AIC
aic_plot_pbmc <- ggplot(mse_df_pbmc, aes(x = num_bins, y = aic
)) +
  geom_point() +
  geom_line() +
  # geom_vline(xintercept = 256, linetype="dashed", 
  #               color = "red", size=0.5) +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#377eb8")) +
  ylab("AIC") +
  xlab("total number of bins")
## Effective number of bins along x-axis

mse_plot_pbmc <- ggplot(mse_df_pbmc, aes(x = num_bins,
                                       y = mse
)) +
  geom_point() +
  geom_line() +
   # geom_vline(xintercept = 256, linetype="dashed", 
   #              color = "red", size=0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#377eb8")) +
  ylab("MSE") +
  xlab("total number of bins")
